---
title: "Project 2.1 - Comparison of Level and Observed Coverage: Gaussian CI for 𝜇"
author: "Duy My Phuong Doan"
date: "`r Sys.Date()`"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
library(shiny)
knitr::opts_chunk$set(echo = TRUE, fig.width = 7, fig.height = 4.5)
set.seed(2024)
```

## 1. Goal

We study the **observed coverage** of the confidence interval (CI) for the mean
$\mu$ of a Gaussian population $X \sim N(\mu, \sigma^2)$, comparing two cases:

- **Case A**: $\sigma^2$ **known** &nbsp;→ CI based on the Normal (Z) pivot
- **Case B**: $\sigma^2$ **unknown** &nbsp;→ CI based on the Student's t pivot

For a nominal confidence level $1-\alpha$, we check whether the *proportion*
of simulated intervals that actually contain the true $\mu$ matches the
nominal level, for different sample sizes $n$.

---

## 2. Theoretical CIs

**Case A (sigma known):**
$$\bar X \pm z_{\alpha/2}\,\frac{\sigma}{\sqrt{n}}$$

**Case B (sigma unknown):**
$$\bar X \pm t_{\alpha/2,\,n-1}\,\frac{S}{\sqrt{n}}$$

where $S^2$ is the sample variance, computed directly from the data
(using only the sample, no knowledge of the true $\sigma$):

$$S^2 = \frac{1}{n-1}\sum_{i=1}^{n}(X_i - \bar X)^2, \qquad S = \sqrt{S^2}$$

Dividing by $n-1$ (Bessel's correction) makes $S^2$ an **unbiased**
estimator of $\sigma^2$, i.e. $E(S^2) = \sigma^2$. Note that $S$ itself is
*not* an unbiased estimator of $\sigma$ (only $S^2$ is unbiased for
$\sigma^2$) — a common subtlety worth remembering. In R, `sd(x)` computes
exactly this formula, matching the `s <- sd(x)` line used in the
simulation function below.

---

## 3. Theory: what does "exact CI" mean?

A confidence interval is called **exact** when its true coverage probability
equals the nominal level $1-\alpha$ **for every sample size $n$**, not just
in the limit as $n \to \infty$. This is in contrast to an **asymptotic
(approximate) CI**, whose coverage only approaches $1-\alpha$ as $n$ grows
large, and can be quite far off from the nominal level for small $n$.

**Case A — $\sigma^2$ known.** The sample mean is *exactly* Gaussian for
every $n$ (not just approximately, and not only for large $n$):
$$\bar X \sim N\!\left(\mu, \frac{\sigma^2}{n}\right) \quad\Longrightarrow\quad
Z = \frac{\bar X - \mu}{\sigma/\sqrt{n}} \sim N(0,1) \ \text{ exactly, for every } n$$

Since $Z$'s distribution never depends on $\mu$ or $\sigma$ (it is a
**pivotal quantity**), we can invert $P(-z_{\alpha/2} \le Z \le z_{\alpha/2}) = 1-\alpha$
to get a CI whose coverage is *exactly* $1-\alpha$, regardless of $n$.

**Case B — $\sigma^2$ unknown.** Replacing $\sigma$ with the sample standard
deviation $S$ changes the pivotal quantity's distribution — but it is a
**known, exact** distribution, not an approximation:
$$T = \frac{\bar X - \mu}{S/\sqrt{n}} \sim t_{n-1} \ \text{ exactly, for every } n \ge 2$$

This is a classical result (originally due to Gosset/"Student", 1908): $T$
follows a Student's t distribution with $n-1$ degrees of freedom **exactly**,
again for any sample size — not just asymptotically.

**Why this matters.** Many CIs used in practice (e.g. for Poisson $\lambda$,
or MLE-based CIs in general) rely on the **Central Limit Theorem** and are
only *asymptotically* valid — their coverage can be noticeably below the
nominal level when $n$ is small. The two Gaussian CIs above are a rare case
where the exact finite-sample distribution of the pivot is known in closed
form, so there is **no need to wait for large $n$**: coverage is correct
even at $n=2$ or $n=3$. Sections 4-6 verify this with a fixed simulation
study.

**References for this section.**

- Casella, G. and Berger, R. L. (2002). *Statistical Inference* (2nd ed.),
  Duxbury. Theorem 5.3.1 (distribution of the sample mean and variance
  for a Normal population) is the basis for both exact-CI derivations
  above, including the independence of $\bar X$ and $S^2$ used to derive
  the distribution of $T$.
- Cochran, W. G. (1934). "The distribution of quadratic forms in a normal
  system, with applications to the analysis of covariance." *Mathematical
  Proceedings of the Cambridge Philosophical Society*, 30(2), 178-191.
  (Establishes the independence of $\bar X$ and $S^2$ under normality,
  used implicitly in deriving the exact distribution of $T$.)
- Student [Gosset, W. S.] (1908). "The probable error of a mean."
  *Biometrika*, 6(1), 1-25. (Original derivation of the t distribution.)

---

## 4. Simulation function

This function is used for the fixed-parameter study below (Sections 5 and 5b).

```{r sim-function}
# Simulate M samples of size n from N(mu, sigma^2) and compute
# observed coverage AND average width of the two CIs (sigma known / unknown)
simulate_coverage <- function(M = 10000, n, mu = 0, sigma = 1, alpha = 0.05) {
  z_crit <- qnorm(1 - alpha / 2)
  t_crit <- qt(1 - alpha / 2, df = n - 1)

  cover_known   <- logical(M)
  cover_unknown <- logical(M)
  width_known_vec   <- numeric(M)
  width_unknown_vec <- numeric(M)

  for (m in 1:M) {
    x <- rnorm(n, mean = mu, sd = sigma)
    xbar <- mean(x)
    s    <- sd(x)   # sample sd, denominator n-1

    # Case A: sigma known
    margin_A <- z_crit * sigma / sqrt(n)
    cover_known[m] <- (mu >= xbar - margin_A) && (mu <= xbar + margin_A)
    width_known_vec[m] <- 2 * margin_A

    # Case B: sigma unknown
    margin_B <- t_crit * s / sqrt(n)
    cover_unknown[m] <- (mu >= xbar - margin_B) && (mu <= xbar + margin_B)
    width_unknown_vec[m] <- 2 * margin_B
  }

  c(
    n = n,
    mu = mu,
    sigma = sigma,
    coverage_known    = mean(cover_known),
    coverage_unknown  = mean(cover_unknown),
    width_known       = mean(width_known_vec),
    width_unknown     = mean(width_unknown_vec)
  )
}
```

---

## 5. Run simulations for different sample sizes (fixed study)

```{r run-sim}
M <- 10000
mu <- 5
sigma <- 2
alpha <- 0.05
n_values <- c(3, 5, 10, 20, 30, 50, 100)

results <- t(sapply(n_values, function(n) {
  simulate_coverage(M = M, n = n, mu = mu, sigma = sigma, alpha = alpha)
}))
results <- as.data.frame(results)
results
```

```{r table}
knitr::kable(results[, c("n", "coverage_known", "coverage_unknown")],
             col.names = c("n", "Coverage (sigma known)", "Coverage (sigma unknown)"),
             digits = 4,
             caption = paste0("Observed coverage vs nominal level = ", 1 - alpha))
```

```{r plot}
plot(results$n, results$coverage_known, type = "b", pch = 16, col = "blue",
     ylim = c(min(results$coverage_known, results$coverage_unknown) - 0.01, 1),
     xlab = "Sample size (n)", ylab = "Observed coverage",
     main = "Observed coverage of 95% CI for mu")
lines(results$n, results$coverage_unknown, type = "b", pch = 17, col = "red")
abline(h = 1 - alpha, lty = 2, col = "darkgreen")
legend("topright",
       legend = c("sigma known (Z)", "sigma unknown (t)", "nominal level"),
       col = c("blue", "red", "darkgreen"), lty = c(1,1,2), pch = c(16,17,NA))
```

```{r width-table}
knitr::kable(results[, c("n", "width_known", "width_unknown")],
             col.names = c("n", "Avg width (sigma known)", "Avg width (sigma unknown)"),
             digits = 4,
             caption = "Average CI width: known vs unknown sigma")
```

---

## 5b. Varying the true parameter values (mu, sigma), n fixed

Section 5 varied the sample size n while keeping mu and sigma fixed. To
complete step (iii) of the assignment — "different parameter(s) values" —
we now do the reverse: fix n and vary the true mu and sigma, to check
whether coverage really is independent of the parameter values, as
claimed by the pivotal-quantity argument in Section 3.

```{r run-sim-params}
n_fixed <- 20

param_grid <- data.frame(
  mu    = c(-5, 0, 5, 5, 5),
  sigma = c(2, 2, 0.5, 2, 5)
)

param_results <- t(mapply(function(mu, sigma) {
  simulate_coverage(M = M, n = n_fixed, mu = mu, sigma = sigma, alpha = alpha)
}, param_grid$mu, param_grid$sigma))
param_results <- as.data.frame(param_results)
param_results
```

```{r table-params}
knitr::kable(param_results[, c("mu", "sigma", "coverage_known", "coverage_unknown")],
             col.names = c("mu", "sigma", "Coverage (sigma known)", "Coverage (sigma unknown)"),
             digits = 4,
             caption = paste0("Observed coverage at n = ", n_fixed,
                               " for different true (mu, sigma), nominal level = ", 1 - alpha))
```

As expected, coverage stays close to 0.95 regardless of the value of mu
or sigma — confirming that Z and T are genuinely pivotal quantities:
their sampling distribution (and hence the coverage of the CIs built from
them) does not depend on the true parameter values, only on n.

---

## 6. Discussion

- Both CIs are **exact** (not asymptotic), as derived in Section 3, so
  observed coverage stays close to the nominal level $1-\alpha$ for every
  n, including very small n (e.g. n = 3).
- Any deviation seen in the table is just Monte Carlo noise (which shrinks
  as M increases) — try increasing `M` to 50000 to see coverage converge
  even more tightly to `r 1-alpha`.
- The CI width table shows the "unknown sigma" case is noticeably wider at
  small n (the cost of having to estimate sigma from the data), and the gap
  shrinks as n grows.

---

